---
title: "Comparsion of PSSM"
author: "Richard Martin"
format:
  html:
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
---

```{r, include=FALSE}
library(tidyverse)
library(here)
library(readxl)
library(conflicted)
library(janitor)
library(plotly)
library(qdapRegex) #for ex_between
conflicts_prefer(dplyr::filter)
```

```{r, message=FALSE, warning=FALSE}
#constants----------------------
letters <- "[:alpha:]"
#functions---------------------------
apply_props <- function(tbbl, val){
  tbbl|>
    mutate(count=prop*val)
}

#compare current and previous pssm in terms of occupation proportions.

noc_mapping <- read_csv(here("data","noc_mapping.csv"))|>
  clean_names()|>
  select(noc_2016=noc_2016_v1_3_code, noc_2021=noc_2021_v1_0_code)

old_noc_props <- read_excel(here("data",
                            "B.C. Post-Secondary Supply Model 2019-20 to 2030-31 2021-09-21 Internal Use.xlsx"),
                       sheet = "Occupation Projections",
                       na = "N")|>
  filter(`Region Name`=="British Columbia",
         `Age Group`=="17 to 29"
         )|>
  pivot_longer(cols=starts_with("2"))|>
  clean_names()|>
  mutate(noc_2016=as.numeric(noc_2016))|>
  full_join(noc_mapping)|>
  group_by(noc_2021)|>
  summarize(old_value=sum(value, na.rm = TRUE))|>
  ungroup()|>
  mutate(old_prop=old_value/sum(old_value))



new_noc_props <- read_excel(here("data",
                                 #"draft_internal_use_PSSM_2023-24_to_2034-35_20241002.xlsx"),
                                 "draft_internal_use_PSSM_2023-24_to_2034-35_20241018.xlsx"),
                            sheet = "Occupation Projections",
                            na = "N",
                            skip=1)|>
  filter(`Region Name`=="British Columbia",
         `Age Group`=="17 to 29")|>
  pivot_longer(cols=starts_with("2"))|>
  clean_names()|>
  group_by(noc_2021, occupation_description)|>
  summarize(new_value=sum(value, na.rm = TRUE))|>
  ungroup()|>
  mutate(new_prop=new_value/sum(new_value),
         noc_2021=as.numeric(noc_2021))

noc_props <- inner_join(new_noc_props, old_noc_props, by="noc_2021")

new_vs_old_plt <- ggplot(noc_props, aes(old_prop,
                             new_prop,
                             text=paste0(
                               "Occupation: ",
                               occupation_description,
                               "\n Previous Proportion= ",
                               scales::percent(old_prop, accuracy = .001),
                               "\n Current Proportion= ",
                               scales::percent(new_prop, accuracy = .001)
                               )
                             )
              )+
  geom_abline(slope = 1, intercept = 0, col="white", lwd=2)+
  geom_point(alpha=.25)+
  scale_x_continuous(trans="log10", labels = scales::percent)+
  scale_y_continuous(trans="log10", labels=scales::percent)+
  labs(x="Proportion of BC Grads in Occupation: Previous PSSM",
       y="Proportion of BC Grads in Occupation: Current PSSM",
       title="Comparison of PSSM BC graduate proportions by Occupation"
       )

#compare with props based on bc grad counts (by cip) and canada cip-noc table.

#read in the data-------------------------------------

cip_counts_col_names <- unlist(as.vector(read_csv(here("data","cip_counts_bc.csv"), skip=12, n_max = 1, col_names = FALSE)))

cip_counts <- read_csv(here("data","cip_counts_bc.csv"), skip=14, na = "..", col_names = cip_counts_col_names)|>
  mutate(CIP=ex_between(`Field of study 5`,"[","]"),
         CIP=str_remove_all(CIP, letters),
         CIP=str_pad(CIP, width=5, side = "left", pad = "0"),
         field_of_study=word(`Field of study 5`, sep="\\["), .before = everything())|>
  select(-`Field of study 5`)|>
  pivot_longer(cols = starts_with("2"))|>
  group_by(CIP, field_of_study)|>
  summarise(mean_grads=mean(value, na.rm = TRUE))

cip_noc_long <- vroom::vroom(here("data","cip_2_noc_canada.csv"), skip = 13)[-1,]
colnames(cip_noc_long)[1] <- "field_of_study"
cip_noc_long <- cip_noc_long|>
  janitor::remove_empty("cols")|>
  pivot_longer(cols=-field_of_study, names_to = "noc", values_to = "count")|>
  mutate(count=as.numeric(str_remove_all(count,",")),
         CIP=str_sub(field_of_study, 1, 5),
         field_of_study=str_sub(field_of_study, 7))


richs_supply <- cip_noc_long|>
  group_by(CIP, field_of_study)|>
  mutate(prop=count/sum(count, na.rm = TRUE))|>
  select(-count)|>
  nest()|>
  inner_join(cip_counts, by="CIP")|>
  mutate(data=map2(data, mean_grads, apply_props))|>
  select(data)|>
  unnest(data)|>
  group_by(noc)|>
  summarize(supply_by_occupation=sum(count, na.rm = TRUE))|>
  ungroup()|>
  mutate(richs_prop=supply_by_occupation/sum(supply_by_occupation, na.rm = TRUE),
         noc_2021=as.numeric(str_sub(noc, 1,5))
         )

rich_vs_new <- inner_join(richs_supply, new_noc_props)

rich_vs_new_plt <- ggplot(rich_vs_new, aes(richs_prop,
                             new_prop,
                             text=paste0(
                               "Occupation: ",
                               occupation_description,
                               "\n Rich's Proportion= ",
                               scales::percent(richs_prop, accuracy = .001),
                               "\n PSSM Proportion= ",
                               scales::percent(new_prop, accuracy = .001)
                             )
)
)+
  geom_abline(slope = 1, intercept = 0, col="white", lwd=2)+
  geom_point(alpha=.25)+
  scale_x_continuous(trans="log10", labels = scales::percent)+
  scale_y_continuous(trans="log10", labels=scales::percent)+
  labs(x="Proportion of BC Grads in Occupation: Rich",
       y="Proportion of BC Grads in Occupation: Current PSSM",
       title="Comparison of PSSM vs. applying CIP/NOC proportions to mean CIP Counts"
  )

#Compare with last version of PSSM-------------------------------------
last_version_noc_props <- read_excel(here("data",
                                          "draft_internal_use_PSSM_2023-24_to_2034-35_20241002.xlsx"),
                                    sheet = "Occupation Projections",
                                    na = "N",
                                    skip=1)|>
  filter(`Region Name`=="British Columbia",
         `Age Group`=="17 to 29")|>
  pivot_longer(cols=starts_with("2"))|>
  clean_names()|>
  group_by(noc_2021, occupation_description)|>
  summarize(last_value=sum(value, na.rm = TRUE))|>
  ungroup()|>
  mutate(last_prop=last_value/sum(last_value),
         noc_2021=as.numeric(noc_2021))

version_compare <- inner_join(new_noc_props, last_version_noc_props)

options(scipen = 999)
version_compare_plt <- ggplot(version_compare,  aes(new_value,
                             last_value,
                             text=paste0(
                               "Occupation: ",
                               occupation_description,
                               "\n Last Value= ",
                               scales::comma(last_value),
                               "\n New Value= ",
                               scales::comma(new_value)
                             )
)
)+
  geom_abline(slope = 1, intercept = 0, col="white", lwd=2)+
  geom_jitter(alpha=.25, width=.05, height=.05)+
  scale_x_continuous(trans="log10")+
  scale_y_continuous(trans="log10")+
  labs(x="BC Grads in Occupation: New Version",
       y="BC Grads in Occupation: Last version",
       title="Comparison of 2 versions of current PSSM"
  )
```

## Comparison with previous PSSM

-   Here we compare the by occupation grad proportions for the new vs the old PSSM (for `Age Group`=="17 to 29")

```{r, warning=FALSE}
ggplotly(new_vs_old_plt, tooltip="text")
```


## Comparison with my method

-   My method is to take the historic mean BC grad counts by CIP (Table: 37-10-0183-01), and apply proportions from Canada's CIP-NOC Table: 98-10-0403-01 (for ages 15-24) to form a prediction of where (i.e. NOCs) the grads will end up working. 

```{r, warning=FALSE}
ggplotly(rich_vs_new_plt, tooltip = "text")
```

## Comparison with first draft

```{r, warning=FALSE}
ggplotly(version_compare_plt, tooltip = "text")
```


